---
title: "USDA Census of Agriculture — AFPC Partner CSV Exports (Alaska)"
subtitle: "Table 1 Summary (State + Ag Census Areas) and Trimmed Ch 2 Compact Files"
author: |
  | Michael S. Jones
  | Research Assistant Professor of Economics
  | Institute of Social and Economic Research (ISER)
  | University of Alaska Anchorage
  | msjones6@alaska.edu
  | Detailed prompts with Claude Opus models to generate the following extract code
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}

# ==========================================================================================================================
#SET THE FOLLOWING FOR YOUR MACHINE
# NASS API key — uses NASS_API_KEY from .Renviron when set, otherwise this value
NASS_API_KEY <- Sys.getenv("NASS_API_KEY", unset = "ENTER_YOUR_KEY")

# Output directory — adjust to your local path (Google Drive mapped path for Windows)
folder_prefix <- "ENTER_YOUR_DESIRED_FILE_PATH"
# ==========================================================================================================================




# ============================================================================
# PARTNER EXPORT PIPELINE — USDA Census of Agriculture, Alaska
# ----------------------------------------------------------------------------
# One knit  →  four CSVs in OUT_DIR/partner_exports/
#   (1) Table 1 (State Summary Highlights) — state level, all Census years
#   (2) Table 1 (State Summary Highlights) — Ag Census Area level, all Census years
#   (3) Ch 2 compact file — state (+ US) level, same trim as the original pull
#   (4) Ch 2 compact file — county / Ag Census Area level, same trim as before
#
# Every API call is cached as its own .rds, so re-knitting only re-pulls calls
# that are missing or previously failed. Set REFRESH_CACHE <- TRUE to force a
# full re-pull (NASS revisions, 2027 Census release).
#
# Inputs (same folder as this Rmd):
#   ch2_metadata_lookup.csv                               (required)
#   User-Guide-2022-Census-of-Ag-Data-on-Quick-Stats 1.xlsx (optional; gives
#     Table 1 its published row order — falls back to lookup order if absent)
# ============================================================================

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# ============================================================================
# PACKAGES
# ============================================================================
required_packages <- c("dplyr", "tidyr", "readr", "stringr", "tibble",
                       "glue", "janitor", "rnassqs", "readxl", "knitr")

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, quiet = TRUE, repos = "https://cran.rstudio.com/")
  }
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

# ============================================================================
# CONFIGURATION — EDIT THESE
# ============================================================================


# State
STATE_ALPHA <- "AK"

# # Keep US (NATIONAL) rows in export (3), as the original state compact file did
# PULL_NATIONAL <- TRUE
# Keep US (NATIONAL) rows in export (3), as the original state compact file did
PULL_NATIONAL <- TRUE

# Data source for the four exports
#   "saved" = rebuild from the unfiltered master CSVs the original pulls already
#             wrote to OUT_DIR — no API calls
#   "api"   = pull fresh from Quick Stats (cached per call; use after a new Census release)
#DATA_SOURCE        <- "saved"
DATA_SOURCE        <- "api"

MASTER_STATE_FILE  <- glue("afpc_census_of_ag_{STATE_ALPHA}_all_years.csv")         # state (+ US) master
MASTER_COUNTY_FILE <- glue("afpc_census_of_ag_{STATE_ALPHA}_county_all_years.csv")  # county master
STATE_FIPS         <- "02"   # fills geo_code in export (1); the saved masters don't carry it

OUT_DIR       <- paste0(folder_prefix, "afpc_pull")
EXPORT_SUBDIR <- "partner_exports"      # the four CSVs land in OUT_DIR/partner_exports

# API response cache (lives in OUT_DIR/_nass_api_cache, outside the export folder)
USE_CACHE     <- TRUE

REFRESH_CACHE <- FALSE    # TRUE = wipe this lookup's cache and re-pull everything
#REFRESH_CACHE <- TRUE    # TRUE = wipe this lookup's cache and re-pull everything

# Census years kept in all four exports
CENSUS_YEARS      <- c(1997, 2002, 2007, 2012, 2017, 2022, 2027, 2032, 2037)
CENSUS_YEARS_ONLY <- TRUE

# Table 1 = Ch 2 Table 1, "State Summary Highlights" in the 2022 User Guide
TABLE1_NUM <- 1

# Added rows at the end of exports (1) and (2): food fish (hatchery) context from
# Ch 2 Table 22, and Table 1 sales with food fish netted out
ADD_CONTEXT_ROWS          <- TRUE
FOOD_FISH_MISSING_AS_ZERO <- TRUE   # no food fish row for a geography-year = $0 (none reported)

API_TIMEOUT       <- 120   # seconds before one call is abandoned and split into smaller calls
MAX_CONSEC_FAILS  <- 3     # stop the pull after this many failed calls in a row (re-knit resumes)

# Files (1) and (2) share one schema, so partners can stack them:
# `geo_level, state_alpha, geo_code, geo_name, year, table_row, sector_desc, group_desc, data_item, short_desc,
# domain_desc, domaincat_desc, unit_desc, value, value_num, censored, row_type, var_name, note`.
# `table_row` is the line number in the published Table 1; `data_item` is `short_desc`
# plus the size or sales class where the cell has one; `var_name` is a short
# snake_case name for each row.

# **Added rows (end of files 1 and 2).** NASS's market value of agricultural products
# sold includes food fish (excluding catfish and trout). In Alaska that is almost entirely
# salmon hatchery production, which is roughly as large as all crop sales and hides
# the conventional-agriculture picture. Six rows follow the published cells, marked
# by `row_type`:

# | var_name | row_type | What it is |
# |---|---|---|
# | `food_fish_sales_usd` | Added context | Food fish sales and distribution ($), from Ch 2 Table 22 |
# | `food_fish_operations` | Added context | Operations reporting food fish, from Ch 2 Table 22 |
# | `ag_sales_excl_food_fish_usd` | Derived | Total ag sales minus food fish |
# | `animal_sales_excl_food_fish_usd` | Derived | Livestock, poultry and products (incl. other aquaculture) minus food fish |
# | `food_fish_pct_of_ag_sales` | Derived | Food fish as a percent of total ag sales |
# | `ag_sales_excl_food_fish_per_farm_usd` | Derived | Ag sales excl. food fish per farm, food fish operations excluded from the farm count |

# A derived value shows `(D)` when any of its inputs is withheld. The `note` column
# gives each row's formula. Expenses and net cash income can't be separated from
# hatchery operations with published data, so those Table 1 rows are left as NASS
# reports them.

# Compact-file trim for exports (3) and (4) — identical to the original pulls
COMPACT_DOMAINS_KEEP <- c("TOTAL", "NAICS CLASSIFICATION")
DROP_OTHER_COUNTIES  <- TRUE     # drop "OTHER (COMBINED) COUNTIES" rollups in (2) and (4)

# API pacing
API_DELAY         <- 1.5   # seconds between calls
CHUNK_SIZE_STATE  <- 40    # short_desc values per call, STATE / NATIONAL (avoids 414 URI Too Long)
CHUNK_SIZE_COUNTY <- 30    # short_desc values per call, COUNTY

# Output file names — (3) and (4) keep the names partners already have
FILE_T1_STATE       <- glue("afpc_coa_{STATE_ALPHA}_table1_state_all_years.csv")
FILE_T1_AREA        <- glue("afpc_coa_{STATE_ALPHA}_table1_ag_census_area_all_years.csv")
FILE_COMPACT_STATE  <- glue("afpc_census_of_ag_{STATE_ALPHA}_compact.csv")
FILE_COMPACT_COUNTY <- glue("afpc_census_of_ag_{STATE_ALPHA}_county_compact.csv")

# Contact
CONTACT_NOTE <- "Contact: M. Jones (UAA-ISER) • msjones6@alaska.edu"

# ============================================================================
# PATHS — resolved relative to this Rmd (works for Knit, chunk-by-chunk, or render())
# ============================================================================
rmd_dir <- function() {
  p <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NULL)
  if (!is.null(p)) return(dirname(p))
  if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
    p <- tryCatch(rstudioapi::getSourceEditorContext()$path, error = function(e) "")
    if (nzchar(p)) return(dirname(p))
  }
  getwd()
}
RMD_DIR <- rmd_dir()

find_input <- function(fname, pattern = NULL) {
  dirs <- unique(c(RMD_DIR, getwd()))
  hits <- file.path(dirs, fname)
  hits <- hits[file.exists(hits)]
  if (!length(hits) && !is.null(pattern)) {
    hits <- unlist(lapply(dirs, list.files, pattern = pattern, full.names = TRUE))
  }
  if (length(hits)) hits[1] else NA_character_
}

LOOKUP_PATH <- find_input("ch2_metadata_lookup.csv")
GUIDE_PATH  <- find_input("User-Guide-2022-Census-of-Ag-Data-on-Quick-Stats 1.xlsx",
                          pattern = "^User-Guide-2022-Census-of-Ag.*\\.xlsx$")

# Fallback for non-Windows / testing (same rule as the original pulls)
if (!dir.exists(dirname(OUT_DIR))) OUT_DIR <- file.path(RMD_DIR, "afpc_pull")
EXPORT_DIR <- file.path(OUT_DIR, EXPORT_SUBDIR)
dir.create(EXPORT_DIR, recursive = TRUE, showWarnings = FALSE)

cat("========================================\n")
cat("AFPC Census of Ag — Partner Exports —", STATE_ALPHA, "\n")
cat("Started:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
cat("Exports:", EXPORT_DIR, "\n")
cat("========================================\n\n")

```

# Overview

This workbook produces the four partner CSVs from a single, cached pull of the
NASS Quick Stats API (`rnassqs`), driven by the same Ch 2 metadata crosswalk
(`ch2_metadata_lookup.csv`) used for the original AFPC state and county pulls.

| # | File | Geography | Contents |
|---|------|-----------|----------|
| 1 | `r FILE_T1_STATE` | State (`r STATE_ALPHA`) | Table 1 *State Summary Highlights* — every Table 1 cell NASS publishes for the state, all Census years, tidy long format in published row order |
| 2 | `r FILE_T1_AREA` | Ag Census Areas | Same Table 1 cells and layout, one block per Ag Census Area |
| 3 | `r FILE_COMPACT_STATE` | State (+ US) | Ch 2 Tables 1–57, same trim as before (Census years; `TOTAL` + `NAICS CLASSIFICATION` domains) |
| 4 | `r FILE_COMPACT_COUNTY` | Ag Census Areas | Ch 2 Tables 1–57 at county level, same trim as before (`OTHER (COMBINED)` rollups dropped) |

Files (1) and (2) share one schema, so partners can stack them:
`geo_level, state_alpha, geo_code, geo_name, year, table_row, sector_desc, group_desc, data_item, short_desc,
domain_desc, domaincat_desc, unit_desc, value, value_num, censored`.
`table_row` is the line number in the published Table 1; `data_item` is `short_desc`
plus the size or sales class where the cell has one.

**Why Table 1 is built from the full pull, not the compact files.** Table 1 includes
farms by size class (`AREA OPERATED` domain) and farms by value of sales
(`FARM SALES` domain). The compact files' domain whitelist drops those rows, so
files (1) and (2) are matched cell-by-cell on `short_desc` × `domain_desc` ×
`domaincat_desc` against the raw pull instead.

**Ag Census Areas.** For Alaska, Quick Stats' `COUNTY` level is NASS's five
Agriculture Census Areas — Aleutian Islands, Anchorage, Fairbanks, Juneau and
Kenai Peninsula — each grouping several boroughs/census areas for data collection.

# Load Metadata Lookup

```{r load_lookup}
stopifnot("ch2_metadata_lookup.csv not found — place it next to this Rmd" = !is.na(LOOKUP_PATH))

lookup_raw <- read_csv(LOOKUP_PATH, show_col_types = FALSE) %>%
  clean_names()

target_short_descs <- unique(lookup_raw$short_desc)

cat("Lookup:", LOOKUP_PATH, "\n")
cat("  Rows:", nrow(lookup_raw), "| Tables:", n_distinct(lookup_raw$table),
    "| Unique short_desc:", length(target_short_descs), "\n")

# ------------------------------------------------------------------
# Batch plan: one API batch per sector_desc × group_desc
# (deterministic order so cached chunk files line up across runs)
# ------------------------------------------------------------------
batch_plan <- lookup_raw %>%
  distinct(sector_desc, group_desc, short_desc) %>%
  group_by(sector_desc, group_desc) %>%
  summarise(n_items = n(), items = list(sort(unique(short_desc))), .groups = "drop") %>%
  arrange(desc(n_items), sector_desc, group_desc) %>%
  mutate(batch_id = row_number())

# ------------------------------------------------------------------
# short_desc → Ch 2 table tag (same rule as the original pulls)
# ------------------------------------------------------------------
sd_to_tables <- lookup_raw %>%
  distinct(short_desc, table, table_name) %>%
  group_by(short_desc) %>%
  summarise(
    ch2_tables     = paste(sort(unique(table)), collapse = ";"),
    ch2_table_name = first(table_name),
    .groups        = "drop"
  )

# ------------------------------------------------------------------
# API cache folder — keyed to state + lookup contents, so editing the
# lookup automatically starts a fresh cache
# ------------------------------------------------------------------
lookup_sig <- substr(unname(tools::md5sum(LOOKUP_PATH)), 1, 8)
CACHE_DIR  <- file.path(OUT_DIR, "_nass_api_cache", glue("{STATE_ALPHA}_{lookup_sig}"))
if (REFRESH_CACHE && dir.exists(CACHE_DIR)) unlink(CACHE_DIR, recursive = TRUE)
dir.create(CACHE_DIR, recursive = TRUE, showWarnings = FALSE)
cat("  API cache:", CACHE_DIR, "\n")
```

```{r table1_spec}
# ============================================================================
# TABLE 1 CELL SPECIFICATION — one row per published Table 1 cell
# ============================================================================
t1_spec <- lookup_raw %>%
  filter(table == TABLE1_NUM) %>%
  distinct(table_name, short_desc, domain_desc, domaincat_desc, unit_desc)

# Published row order ("Census Row") from the User Guide workbook, when available
t1_rows <- tibble(short_desc = character(), domain_desc = character(),
                  domaincat_desc = character(), table_row = integer())

if (!is.na(GUIDE_PATH)) {
  t1_rows <- tryCatch({
    g <- readxl::read_excel(GUIDE_PATH, sheet = "Metadata - Ch 2 Tables 1 to 57")
    names(g) <- janitor::make_clean_names(sub("\n.*$", "", names(g)))
    g %>%
      filter(census_table == TABLE1_NUM) %>%
      transmute(short_desc, domain_desc, domaincat_desc,
                table_row = as.integer(census_row)) %>%
      distinct()
  }, error = function(e) {
    cat("Could not read row order from the User Guide:", conditionMessage(e), "\n")
    t1_rows
  })
}

t1_spec <- t1_spec %>%
  left_join(t1_rows, by = c("short_desc", "domain_desc", "domaincat_desc"))

# Any cell without a published row number goes to the end, in lookup order
if (anyNA(t1_spec$table_row)) {
  start <- max(c(0L, t1_spec$table_row), na.rm = TRUE)
  t1_spec$table_row[is.na(t1_spec$table_row)] <- start + seq_len(sum(is.na(t1_spec$table_row)))
}

t1_spec <- t1_spec %>%
  mutate(data_item = if_else(domain_desc == "TOTAL",
                             short_desc,
                             paste0(short_desc, " | ", domaincat_desc))) %>%
  arrange(table_row)

stopifnot("Table 1 cells are not unique in the lookup" =
            !anyDuplicated(t1_spec[c("short_desc", "domain_desc", "domaincat_desc")]))

cat("Table 1:", unique(t1_spec$table_name), "\n")
cat("  Cells:", nrow(t1_spec), "| Unique short_desc:", n_distinct(t1_spec$short_desc), "\n")
cat("  Row order:", if (nrow(t1_rows)) "published (User Guide)" else "lookup order (User Guide not found)", "\n")
```

# API Pull (Cached)

```{r api_engine}
# ============================================================================
nassqs_auth(key = NASS_API_KEY)

# Defaults, in case the setup chunk predates these settings
if (!exists("API_TIMEOUT"))      API_TIMEOUT      <- 120  # seconds before one call is abandoned and split
if (!exists("MAX_CONSEC_FAILS")) MAX_CONSEC_FAILS <- 3    # failed calls in a row before the pull stops

MIN_SPLIT    <- 5   # smallest chunk a timed-out request is split down to
consec_fails <- 0   # circuit-breaker counter, reset by any successful call

api_log <- list()
log_call <- function(geo, label, status, rows = NA_integer_, secs = NA_real_, error = NA_character_) {
  api_log[[length(api_log) + 1]] <<- tibble(geo = geo, call = as.character(label), status = status,
                                            rows = rows, secs = secs, error = error)
}

is_too_big <- function(msg) grepl("exceed|too large|50,?000|\\b413\\b|\\b414\\b|uri too long",
                                  msg, ignore.case = TRUE)
is_timeout <- function(msg) grepl("\\b504\\b|gateway time|timeout was reached|timed out",
                                  msg, ignore.case = TRUE)
is_no_data <- function(msg) grepl("invalid query|bad request|no data|no records",
                                  msg, ignore.case = TRUE)

nass_call <- function(params) {
  res <- tryCatch({
    # as_numeric = FALSE keeps NASS codes — (D), (Z) — instead of turning them into NA
    dat <- if (requireNamespace("httr", quietly = TRUE)) {
      httr::with_config(httr::timeout(API_TIMEOUT), nassqs(params, as_numeric = FALSE))
    } else {
      nassqs(params, as_numeric = FALSE)
    }
    dat %>%
      as_tibble() %>%
      clean_names() %>%
      mutate(across(everything(), as.character))   # uniform types across chunks
  }, error = function(e) e)

  # Circuit breaker: count server-side failures; any real answer resets the count
  if (inherits(res, "error") &&
      !is_no_data(conditionMessage(res)) && !is_too_big(conditionMessage(res))) {
    consec_fails <<- consec_fails + 1
    if (consec_fails >= MAX_CONSEC_FAILS) {
      stop(glue("NASS Quick Stats failed {consec_fails} calls in a row, so the pull stopped. ",
                "Finished chunks are cached; re-knit later to resume.\n",
                "Last error: {conditionMessage(res)}"), call. = FALSE)
    }
  } else {
    consec_fails <<- 0
  }
  res
}

fetch_items <- function(items, base_params) {
  params <- c(base_params, list(short_desc = items))
  res    <- nass_call(params)
  msg    <- if (inherits(res, "error")) conditionMessage(res) else ""

  # One retry for other transient failures (502/503, dropped connection)
  if (inherits(res, "error") && !is_too_big(msg) && !is_timeout(msg) && !is_no_data(msg)) {
    Sys.sleep(5)
    res <- nass_call(params)
    msg <- if (inherits(res, "error")) conditionMessage(res) else ""
  }

  if (!inherits(res, "error")) return(list(data = res, status = "ok", error = NA_character_))

  # Oversized or timed out: split in half and pull each half
  floor_n <- if (is_timeout(msg)) MIN_SPLIT else 1
  if ((is_too_big(msg) || is_timeout(msg)) && length(items) > floor_n) {
    half  <- ceiling(length(items) / 2)
    left  <- fetch_items(items[seq_len(half)], base_params)
    Sys.sleep(API_DELAY)
    right <- fetch_items(items[-seq_len(half)], base_params)
    errs  <- na.omit(c(left$error, right$error))
    return(list(
      data   = bind_rows(left$data, right$data),
      status = if (any(c(left$status, right$status) == "failed")) "failed" else "ok (split)",
      error  = if (length(errs)) paste(errs, collapse = " | ") else NA_character_
    ))
  }

  if (is_no_data(msg)) return(list(data = tibble(), status = "no data", error = NA_character_))

  list(data = tibble(), status = "failed", error = msg)
}

pull_geo <- function(agg_level, chunk_size) {
  out <- list()

  for (b in seq_len(nrow(batch_plan))) {
    bp     <- batch_plan[b, ]
    items  <- bp$items[[1]]
    chunks <- split(items, ceiling(seq_along(items) / chunk_size))

    base_params <- list(
      source_desc    = "CENSUS",
      agg_level_desc = agg_level,
      sector_desc    = bp$sector_desc,
      group_desc     = bp$group_desc
    )
    if (agg_level != "NATIONAL") base_params$state_alpha <- STATE_ALPHA

    for (i in seq_along(chunks)) {
      label      <- glue("{bp$sector_desc} | {bp$group_desc} | chunk {i}/{length(chunks)}")
      cache_file <- file.path(CACHE_DIR, sprintf("%s_cs%02d_b%02d_c%02d.rds",
                                                 agg_level, chunk_size, bp$batch_id, i))

      if (USE_CACHE && file.exists(cache_file)) {
        dat <- readRDS(cache_file)
        log_call(agg_level, label, "cached", nrow(dat))
        out[[length(out) + 1]] <- dat
        next
      }

      t0   <- Sys.time()
      res  <- fetch_items(chunks[[i]], base_params)
      secs <- round(as.numeric(difftime(Sys.time(), t0, units = "secs")), 1)
      log_call(agg_level, label, res$status, nrow(res$data), secs, res$error)

      # Progress to the console only (keeps the HTML report clean)
      cat(sprintf("  [%-8s] %-55s %-10s %6d rows (%.1fs)\n",
                  agg_level, label, res$status, nrow(res$data), secs), file = stderr())
      if (res$status == "failed") {
        cat("             error:", substr(gsub("\\s+", " ", res$error), 1, 160), "\n", file = stderr())
      }

      if (res$status != "failed") {
        if (USE_CACHE) saveRDS(res$data, cache_file)
        out[[length(out) + 1]] <- res$data
      }
      Sys.sleep(API_DELAY)
    }
  }

  bind_rows(out)
}
```

```{r execute_pulls, cache=FALSE}
api_log_df   <- tibble(geo = character(), call = character(), status = character(),
                       rows = integer(), secs = numeric(), error = character())
pull_minutes <- 0

if (DATA_SOURCE == "saved") {
  # ------------------------------------------------------------------
  # Rebuild from the unfiltered masters the original pulls saved.
  # Read everything as text so (D)/(Z) codes and leading-zero codes survive;
  # value_num, censored and the Ch 2 tags are re-derived in the next chunk.
  # ------------------------------------------------------------------
  read_master <- function(fname) {
    path <- file.path(OUT_DIR, fname)
    if (!file.exists(path)) {
      stop(glue("Saved master not found: {path}\n",
                "Point MASTER_*_FILE at it, or set DATA_SOURCE <- \"api\"."), call. = FALSE)
    }
    cat("Reading", path, "\n")
    read_csv(path, col_types = cols(.default = col_character()))
  }

  master_state_raw <- read_master(MASTER_STATE_FILE)
  if (!"state_fips_code" %in% names(master_state_raw)) {
    master_state_raw$state_fips_code <- if_else(master_state_raw$agg_level_desc == "NATIONAL",
                                                 "99", STATE_FIPS)
  }

  raw_state  <- master_state_raw %>% filter(agg_level_desc == "STATE")
  raw_nat    <- if (PULL_NATIONAL) filter(master_state_raw, agg_level_desc == "NATIONAL") else tibble()
  raw_county <- read_master(MASTER_COUNTY_FILE)

  cat("Source: saved master CSVs (no API calls)\n")

} else {
  t_start <- Sys.time()

  raw_state  <- pull_geo("STATE",  CHUNK_SIZE_STATE)
  raw_nat    <- if (PULL_NATIONAL) pull_geo("NATIONAL", CHUNK_SIZE_STATE) else tibble()
  raw_county <- pull_geo("COUNTY", CHUNK_SIZE_COUNTY)

  pull_minutes <- round(as.numeric(difftime(Sys.time(), t_start, units = "mins")), 1)
  api_log_df   <- bind_rows(api_log)
  cat("Source: Quick Stats API, pull finished in", pull_minutes, "minutes\n")
}

cat("  Raw rows — STATE:", nrow(raw_state), "| NATIONAL:", nrow(raw_nat),
    "| COUNTY:", nrow(raw_county), "\n")

if (nrow(api_log_df) > 0) {
  api_log_df %>%
    count(geo, status) %>%
    pivot_wider(names_from = status, values_from = n, values_fill = 0) %>%
    kable(caption = "API calls by geography and outcome ('cached' = reused from a previous knit)")
}
```

# Assemble Master Data

```{r standardize}

# ============================================================================
# STANDARDIZE RAW PULLS
#   value      — published value, whitespace trimmed (e.g. "1,234" or "(D)")
#   value_num  — numeric value; NA when NASS publishes a code instead of a number
#   censored   — TRUE when the published value is a NASS code — (D) withheld,
#                (Z) less than half the unit, (NA), (X), etc. — rather than a number
#
# Every Quick Stats row carries either a number or a code, so a row with no
# number is a code. rnassqs::nassqs() defaults to as_numeric = TRUE, which turns
# (D)/(Z) into NA; the original masters were pulled that way. Pull with
# as_numeric = FALSE (api_engine) to keep NASS's exact code. On older pulls the
# code is taken from CV (%) where NASS repeats it; otherwise value stays blank
# (censored, code unknown) and a warning says how many rows need a re-pull.
# ============================================================================
keep_cols <- c(
  "source_desc", "sector_desc", "group_desc", "commodity_desc",
  "class_desc", "prodn_practice_desc",
  "statisticcat_desc", "domain_desc", "domaincat_desc",
  "short_desc", "unit_desc",
  "agg_level_desc", "state_alpha", "state_name", "state_fips_code",
  "county_code", "county_ansi", "county_name", "asd_code", "asd_desc",
  "year", "freq_desc", "reference_period_desc",
  "value", "cv_percent"
)

standardize <- function(raw) {
  if (nrow(raw) == 0) return(tibble())
  if (!"cv_percent" %in% names(raw)) raw$cv_percent <- NA_character_
  out <- raw %>%
    select(any_of(keep_cols)) %>%
    distinct() %>%
    mutate(
      value      = str_squish(value),
      year       = as.integer(year),
      value_num  = suppressWarnings(parse_number(value)),
      censored   = is.na(value_num),
      value      = if_else(censored,
                           coalesce(str_extract(value, "\\([A-Z]+\\)"),
                                    str_extract(str_squish(cv_percent), "\\((D|Z)\\)")),
                           value)
    ) %>%
    left_join(sd_to_tables, by = "short_desc") %>%
    mutate(
      ch2_tables     = coalesce(ch2_tables, "EXTRA"),
      ch2_table_name = coalesce(ch2_table_name, "(not in Ch2 lookup)")
    )
  n_lost <- sum(out$censored & is.na(out$value))
  if (n_lost > 0) {   # cat, not warning(): the setup chunk hides warnings
    cat("WARNING:", n_lost, "rows have no number and no NASS code (lost upstream).",
        "Re-pull with DATA_SOURCE <- \"api\" to recover the exact codes.\n")
  }
  out
}

keep_years <- function(df) {
  if (CENSUS_YEARS_ONLY) filter(df, year %in% CENSUS_YEARS) else df
}

drop_other <- function(df) {
  if (DROP_OTHER_COUNTIES) filter(df, !grepl("^OTHER", county_name, ignore.case = TRUE)) else df
}

# "ANCHORAGE" → "Anchorage Area" (NASS's published Ag Census Area names)
area_label <- function(x) {
  x <- str_to_title(x)
  if_else(grepl("Area$", x), x, paste(x, "Area"))
}

master_state  <- bind_rows(standardize(raw_state), standardize(raw_nat))
master_county <- standardize(raw_county)

cat("Master STATE/US:", nrow(master_state), "rows |",
    n_distinct(master_state$short_desc), "items | years:",
    paste(sort(unique(master_state$year)), collapse = ", "), "|",
    sum(master_state$censored), "published as a code\n")
cat("Master COUNTY:  ", nrow(master_county), "rows |",
    n_distinct(master_county$short_desc), "items | areas:",
    paste(sort(unique(master_county$county_name)), collapse = ", "), "|",
    sum(master_county$censored), "published as a code\n")

```

# Write Partner Exports

```{r export_helper}
# export_manifest <- list()

# write_export <- function(df, fname, contents) {
#   path <- file.path(EXPORT_DIR, fname)
#   write_csv(df, path)
#   export_manifest[[length(export_manifest) + 1]] <<- tibble(
#     file       = as.character(fname),
#     contents   = contents,
#     rows       = nrow(df),
#     data_items = if (nrow(df)) n_distinct(df$short_desc) else 0L,
#     years      = if (nrow(df)) paste(sort(unique(df$year)), collapse = ", ") else "",
#     size_kb    = round(file.size(path) / 1024)
#   )
#   invisible(path)
# }

# # Match raw rows to the exact published Table 1 cells
# t1_match <- function(df) {
#   df %>%
#     inner_join(t1_spec %>% select(short_desc, domain_desc, domaincat_desc, table_row, data_item),
#                by = c("short_desc", "domain_desc", "domaincat_desc"))
# }
export_manifest <- list()

write_export <- function(df, fname, contents) {
  path <- file.path(EXPORT_DIR, fname)
  write_csv(df, path)
  export_manifest[[length(export_manifest) + 1]] <<- tibble(
    file       = as.character(fname),
    contents   = contents,
    rows       = nrow(df),
    data_items = if (nrow(df)) n_distinct(df$short_desc) else 0L,
    years      = if (nrow(df)) paste(sort(unique(df$year)), collapse = ", ") else "",
    size_kb    = round(file.size(path) / 1024)
  )
  invisible(path)
}

# Table 1 files use the compact files' field format: value and value_num are plain
# numbers (NA where there is no number); the NASS code for those cells goes in nass_code
partner_format <- function(df) {
  df %>%
    mutate(nass_code = if_else(censored, value, NA_character_), .after = censored) %>%
    mutate(value = value_num)
}

# Match raw rows to the exact published Table 1 cells
t1_match <- function(df) {
  df %>%
    inner_join(t1_spec %>% select(short_desc, domain_desc, domaincat_desc, table_row, data_item),
               by = c("short_desc", "domain_desc", "domaincat_desc"))
}
```

## Table 1 Added Rows (Food Fish)

```{r table1_context_rows}
# ============================================================================
# TABLE 1 ADDED ROWS — appended after the published cells in exports (1) and (2)
# ----------------------------------------------------------------------------
# Market value of ag products sold includes food fish (excl. catfish & trout),
# which in Alaska is mostly salmon hatchery production. These rows bring food
# fish in from Ch 2 Table 22 and net it out of the Table 1 sales totals.
#
# Published values only — nothing is estimated:
#   - food fish rows carry the published number or NASS code as-is
#   - a derived row is computed only when every input is a published number;
#     otherwise it is (D) if any input is withheld, else (NA) (e.g. an input
#     that is (Z), or food fish that NASS did not publish for that year)
#   - years when NASS did not publish a food fish item for the state are (NA):
#     sales dollars start in 2007 (1997 had no food fish item; 2002 published
#     farm counts only), operations start in 2002
#   - in a year the item is published, a geography with no row and no food
#     fish operations is 0 (the published table lists no farms there);
#     FOOD_FISH_MISSING_AS_ZERO <- FALSE makes those (NA) too
# ============================================================================
SD_TOTAL_SALES  <- "COMMODITY TOTALS - SALES, MEASURED IN $"
SD_ANIMAL_SALES <- "ANIMAL TOTALS, INCL PRODUCTS - SALES, MEASURED IN $"
SD_FARMS        <- "FARM OPERATIONS - NUMBER OF OPERATIONS"
SD_FISH_SALES   <- "FOOD FISH, (EXCL CATFISH & TROUT) - SALES & DISTRIBUTION, MEASURED IN $"
SD_FISH_OPS     <- "FOOD FISH, (EXCL CATFISH & TROUT) - OPERATIONS WITH SALES & DISTRIBUTION"

# Census years in which NASS published each food fish item for the state
item_years <- function(sd) {
  master_state %>%
    filter(agg_level_desc == "STATE", domain_desc == "TOTAL", short_desc == sd) %>%
    pull(year) %>%
    unique()
}
FISH_SALES_YEARS <- item_years(SD_FISH_SALES)
FISH_OPS_YEARS   <- item_years(SD_FISH_OPS)
cat("Food fish sales published for:", paste(sort(FISH_SALES_YEARS), collapse = ", "),
    "| operations:", paste(sort(FISH_OPS_YEARS), collapse = ", "), "\n")

# Short snake_case names for every published Table 1 cell
# ("COMMODITY TOTALS - SALES, MEASURED IN $ / OPERATION" → commodity_totals_sales_usd_per_operation)
make_var_name <- function(x) {
  x %>%
    str_replace_all("\\$", " usd ") %>%
    str_replace_all("/", " per ") %>%
    str_replace_all("&", " and ") %>%
    str_replace_all("MEASURED IN|\\|", " ") %>%
    janitor::make_clean_names()
}
t1_names <- t1_spec %>% transmute(table_row, var_name = make_var_name(data_item))

# The six added rows, numbered after the last published Table 1 row
derived_spec <- tribble(
  ~var_name, ~data_item, ~short_desc, ~unit_desc, ~row_type, ~note,
  "food_fish_sales_usd",
    "FOOD FISH (HATCHERY), EXCL CATFISH & TROUT - SALES & DISTRIBUTION, MEASURED IN $",
    SD_FISH_SALES, "$", "Added context: Ch 2 Table 22",
    "Food fish (excl. catfish & trout) sales and distribution; in Alaska mostly salmon hatcheries",
  "food_fish_operations",
    "FOOD FISH (HATCHERY), EXCL CATFISH & TROUT - OPERATIONS WITH SALES & DISTRIBUTION",
    SD_FISH_OPS, "OPERATIONS", "Added context: Ch 2 Table 22",
    "Operations reporting food fish (excl. catfish & trout) sales and distribution",
  "ag_sales_excl_food_fish_usd",
    "AG SALES EXCL FOOD FISH - SALES, MEASURED IN $",
    NA, "$", "Derived",
    "commodity_totals_sales_usd minus food_fish_sales_usd",
  "animal_sales_excl_food_fish_usd",
    "ANIMAL SALES EXCL FOOD FISH - SALES, MEASURED IN $",
    NA, "$", "Derived",
    "animal_totals_incl_products_sales_usd minus food_fish_sales_usd (keeps livestock, poultry, products and other aquaculture)",
  "food_fish_pct_of_ag_sales",
    "FOOD FISH - SHARE OF TOTAL AG SALES, MEASURED IN PCT",
    NA, "PCT", "Derived",
    "food_fish_sales_usd / commodity_totals_sales_usd x 100",
  "ag_sales_excl_food_fish_per_farm_usd",
    "AG SALES EXCL FOOD FISH - SALES, MEASURED IN $ / OPERATION",
    NA, "$ / OPERATION", "Derived",
    "ag_sales_excl_food_fish_usd / (farm_operations_number_of_operations minus food_fish_operations)"
) %>%
  mutate(table_row = max(t1_spec$table_row) + row_number())

# NASS's own sector_desc / group_desc for the added rows. The food fish rows are
# native Ch 2 Table 22 items; a derived row has no NASS category of its own, so
# it takes the category of the published item it is built from.
derived_parent <- c(
  food_fish_sales_usd                  = SD_FISH_SALES,
  food_fish_operations                 = SD_FISH_OPS,
  ag_sales_excl_food_fish_usd          = SD_TOTAL_SALES,
  animal_sales_excl_food_fish_usd      = SD_ANIMAL_SALES,
  food_fish_pct_of_ag_sales            = SD_FISH_SALES,
  ag_sales_excl_food_fish_per_farm_usd = SD_TOTAL_SALES
)
derived_spec <- derived_spec %>%
  mutate(parent_sd = unname(derived_parent[var_name])) %>%
  left_join(lookup_raw %>%
              filter(domain_desc == "TOTAL") %>%
              distinct(short_desc, sector_desc, group_desc) %>%
              distinct(short_desc, .keep_all = TRUE),
            by = c(parent_sd = "short_desc")) %>%
  select(-parent_sd)

stopifnot("An added Table 1 row has no NASS sector/group" =
            !anyNA(derived_spec$sector_desc) && !anyNA(derived_spec$group_desc))

# Note appended to a row when its food fish input was not a published row
status_note <- c(
  none          = "no food fish reported here in a year NASS published food fish, counted as 0",
  missing       = "food fish operations reported but no food fish sales row published",
  not_published = "NASS did not publish this food fish item for this Census year"
)

# Published Table 1 rows, standard export columns
t1_publish_rows <- function(base) {
  base %>%
    t1_match() %>%
    left_join(t1_names, by = "table_row") %>%
    transmute(geo_level, state_alpha, geo_code, geo_name,
              year, table_row, sector_desc, group_desc, data_item,
              short_desc, domain_desc, domaincat_desc, unit_desc,
              value, value_num, censored,
              row_type = "NASS Table 1", var_name, note = NA_character_)
}

# Added food fish rows for every geography-year that has a total-sales row
t1_context_rows <- function(base) {
  if (!ADD_CONTEXT_ROWS) return(tibble())
  keys  <- c("geo_level", "state_alpha", "geo_code", "geo_name", "year")
  parts <- c(total = SD_TOTAL_SALES, animal = SD_ANIMAL_SALES, farms = SD_FARMS,
             fish = SD_FISH_SALES, fish_ops = SD_FISH_OPS)

  comp <- base %>%
    filter(domain_desc == "TOTAL", short_desc %in% parts) %>%
    mutate(part = names(parts)[match(short_desc, parts)]) %>%
    distinct(geo_level, state_alpha, geo_code, geo_name, year, part, .keep_all = TRUE) %>%
    select(all_of(keys), part, value, value_num, censored) %>%
    pivot_wider(names_from = part, values_from = c(value, value_num, censored), names_sep = ".")

  for (p in names(parts)) {     # make sure every input column exists (NA = no row)
    for (col in c("value", "value_num", "censored")) {
      nm <- paste0(col, ".", p)
      if (!nm %in% names(comp)) comp[[nm]] <- if (col == "value_num") NA_real_ else NA
    }
  }

  comp <- comp %>%
    filter(!is.na(value_num.total) | censored.total %in% TRUE)
  if (nrow(comp) == 0) return(tibble())

  # Status of the two food fish inputs: published row / none (0) / missing / not published
  comp <- comp %>%
    mutate(
      has_fish_ops = censored.fish_ops %in% TRUE | coalesce(value_num.fish_ops > 0, FALSE),
      fish_status = case_when(
        !is.na(censored.fish)                                 ~ "published",
        !year %in% FISH_SALES_YEARS                           ~ "not_published",
        !has_fish_ops & FOOD_FISH_MISSING_AS_ZERO             ~ "none",
        TRUE                                                  ~ "missing"
      ),
      ops_status = case_when(
        !is.na(censored.fish_ops)                             ~ "published",
        !year %in% FISH_OPS_YEARS                             ~ "not_published",
        FOOD_FISH_MISSING_AS_ZERO                             ~ "none",
        TRUE                                                  ~ "missing"
      )
    )
  for (p in c("fish", "fish_ops")) {
    st   <- comp[[if (p == "fish") "fish_status" else "ops_status"]]
    none <- st == "none"
    gone <- st %in% c("missing", "not_published")
    comp[[paste0("value_num.", p)]][none] <- 0
    comp[[paste0("censored.", p)]][none]  <- FALSE
    comp[[paste0("value.", p)]][gone]     <- "(NA)"
    comp[[paste0("censored.", p)]][gone]  <- TRUE
  }

  v <- function(p) comp[[paste0("value_num.", p)]]
  # NASS code for a value: its own code for a published input; for a derived value,
  # (D) if any input is withheld, else (NA) if any input is not a number
  code <- function(...) {
    ps   <- c(...)
    cens <- Reduce(`|`, lapply(ps, function(p) comp[[paste0("censored.", p)]] %in% TRUE))
    if (length(ps) == 1) return(if_else(cens, coalesce(comp[[paste0("value.", ps)]], "(NA)"), NA_character_))
    anyD <- Reduce(`|`, lapply(ps, function(p) comp[[paste0("censored.", p)]] %in% TRUE &
                                           comp[[paste0("value.", p)]] %in% "(D)"))
    case_when(anyD ~ "(D)", cens ~ "(NA)", TRUE ~ NA_character_)
  }
  calc <- function(var, num, cd) {
    bind_cols(comp[keys], tibble(fish_status = comp$fish_status, ops_status = comp$ops_status,
                                 var_name = var, value_num = num, code = cd))
  }

  bind_rows(
    calc("food_fish_sales_usd",                  v("fish"),                    code("fish")),
    calc("food_fish_operations",                 v("fish_ops"),                code("fish_ops")),
    calc("ag_sales_excl_food_fish_usd",          v("total") - v("fish"),       code("total", "fish")),
    calc("animal_sales_excl_food_fish_usd",      v("animal") - v("fish"),      code("animal", "fish")),
    calc("food_fish_pct_of_ag_sales",            100 * v("fish") / v("total"), code("total", "fish")),
    calc("ag_sales_excl_food_fish_per_farm_usd", (v("total") - v("fish")) /
                                                   (v("farms") - v("fish_ops")), code("total", "fish", "farms", "fish_ops"))
  ) %>%
    mutate(
      code      = if_else(is.na(code) & !is.finite(value_num), "(NA)", code),   # e.g. divide by 0
      censored  = !is.na(code),
      value_num = if_else(censored, NA_real_, value_num)
    ) %>%
    left_join(derived_spec, by = "var_name") %>%
    mutate(
      value = case_when(
        censored           ~ code,
        unit_desc == "PCT" ~ sprintf("%.1f", value_num),
        TRUE               ~ format(round(value_num), big.mark = ",", scientific = FALSE, trim = TRUE)
      ),
      status = if_else(var_name == "food_fish_operations", ops_status, fish_status),
      note   = if_else(status == "published", note, paste0(note, "; ", status_note[status])),
      domain_desc    = "TOTAL",
      domaincat_desc = "NOT SPECIFIED"
    ) %>%
    transmute(geo_level, state_alpha, geo_code, geo_name,
              year, table_row, sector_desc, group_desc, data_item,
              short_desc, domain_desc, domaincat_desc, unit_desc,
              value, value_num, censored,
              row_type, var_name, note) %>%
    arrange(geo_name, table_row, year)
}
```

## Table 1 Market Value Copies

```{r table1_market_value_copies}
# ============================================================================
# MARKET VALUE COPIES — crop and animal sales totals repeated under
# ECONOMICS > INCOME, next to total sales (COMMODITY TOTALS)
# ----------------------------------------------------------------------------
# NASS files total market value sold under ECONOMICS > INCOME, but crop sales
# under CROPS > CROP TOTALS and animal sales under ANIMALS & PRODUCTS > ANIMAL
# TOTALS. A copy of each is added under INCOME so all market value totals sit in
# one group. The originals stay in their native groups.
#
# A copy keeps every value and table_row, var_name and data_item; only
# sector_desc / group_desc, row_type and note change. Filter
# row_type != MV_COPY_ROW_TYPE to drop the copies (e.g. before summing or
# pivoting by var_name).
# ============================================================================
MV_COPY_VARS     <- c("crop_totals_sales_usd",
                      "animal_totals_incl_products_sales_usd",
                      "animal_sales_excl_food_fish_usd")
MV_COPY_ROW_TYPE <- "Copy: market value totals"

# Fail loudly if a var_name is renamed upstream
mv_missing <- setdiff(MV_COPY_VARS, c(t1_names$var_name, derived_spec$var_name))
if (length(mv_missing)) {
  stop("Market value copy rows not found in Table 1: ", paste(mv_missing, collapse = ", "), call. = FALSE)
}

add_market_value_copies <- function(df) {
  copies <- df %>%
    filter(var_name %in% MV_COPY_VARS, row_type != MV_COPY_ROW_TYPE) %>%
    mutate(
      note        = paste0("Copy of row ", table_row, " (", row_type, "; NASS group ",
                           sector_desc, " > ", group_desc, "), repeated under ECONOMICS > INCOME ",
                           "with total sales",
                           if_else(is.na(note), "", paste0("; ", note))),
      sector_desc = "ECONOMICS",
      group_desc  = "INCOME",
      row_type    = MV_COPY_ROW_TYPE
    ) %>%
    arrange(geo_name, table_row, year)
  bind_rows(df, copies)
}
```

## (1) Table 1 — State

``` {r export_1_table1_state}
state_base <- master_state %>%
  filter(agg_level_desc == "STATE", state_alpha == STATE_ALPHA) %>%
  keep_years() %>%
  mutate(geo_level = "STATE", geo_code = state_fips_code, geo_name = str_to_title(state_name))

# Published Table 1 cells, then the added food fish rows, then the market value
# copies under ECONOMICS > INCOME at the end of the file
t1_state <- bind_rows(
  t1_publish_rows(state_base) %>% arrange(table_row, year),
  t1_context_rows(state_base)
) %>%
  add_market_value_copies()

write_export(t1_state, FILE_T1_STATE,
             "Table 1 State Summary Highlights: state, all Census years")

cat("Saved:", FILE_T1_STATE, "—", nrow(t1_state), "rows,",
    n_distinct(t1_state$table_row[t1_state$row_type == "NASS Table 1"]), "of", nrow(t1_spec),
    "Table 1 cells present +",
    sum(!t1_state$row_type %in% c("NASS Table 1", MV_COPY_ROW_TYPE)), "added rows +",
    sum(t1_state$row_type == MV_COPY_ROW_TYPE), "market value copies\n")
```

## (2) Table 1 — Ag Census Areas

```{r export_2_table1_areas}
area_base <- master_county %>%
  filter(agg_level_desc == "COUNTY") %>%
  drop_other() %>%
  keep_years() %>%
  mutate(geo_level = "AG CENSUS AREA", geo_code = county_code, geo_name = area_label(county_name))

# Published Table 1 cells, then the added food fish rows, then the market value
# copies under ECONOMICS > INCOME at the end of the file
t1_area <- bind_rows(
  t1_publish_rows(area_base) %>% arrange(geo_name, table_row, year),
  t1_context_rows(area_base)
) %>%
  add_market_value_copies()

write_export(t1_area, FILE_T1_AREA,
             "Table 1 State Summary Highlights: Ag Census Areas, all Census years")

cat("Saved:", FILE_T1_AREA, "—", nrow(t1_area), "rows |",
    n_distinct(t1_area$geo_name), "areas:", paste(sort(unique(t1_area$geo_name)), collapse = ", "),
    "|", sum(!t1_area$row_type %in% c("NASS Table 1", MV_COPY_ROW_TYPE)), "added rows +",
    sum(t1_area$row_type == MV_COPY_ROW_TYPE), "market value copies\n")
```

## (3) Compact — State (+ US)

```{r export_3_compact_state}
compact_state <- master_state %>%
  mutate(value    = value_num,
         censored = grepl("\\(D\\)|\\(H\\)|\\(Z\\)", cv_percent)) %>%
  select(agg_level_desc, state_alpha,
         year, source_desc, short_desc, unit_desc,
         sector_desc, group_desc, commodity_desc,
         domain_desc, domaincat_desc,
         value, value_num, censored, ch2_tables, ch2_table_name) %>%
  keep_years()

if (!is.null(COMPACT_DOMAINS_KEEP)) {
  compact_state <- compact_state %>% filter(domain_desc %in% COMPACT_DOMAINS_KEEP)
}

compact_state <- compact_state %>% arrange(ch2_tables, short_desc, year)

write_export(compact_state, FILE_COMPACT_STATE,
             "Ch 2 Tables 1-57 compact: state (+ US), Census years, TOTAL + NAICS domains")

cat("Saved:", FILE_COMPACT_STATE, "—", nrow(compact_state), "rows |",
    "geographies:", paste(sort(unique(compact_state$agg_level_desc)), collapse = ", "), "\n")
```

## (4) Compact — Ag Census Areas

```{r export_4_compact_county}
compact_county <- master_county %>%
  mutate(value    = value_num,
         censored = grepl("\\(D\\)|\\(H\\)|\\(Z\\)", cv_percent)) %>%
  select(county_name, county_code,
         year, source_desc, short_desc, unit_desc,
         sector_desc, group_desc, commodity_desc,
         domain_desc, domaincat_desc,
         value, value_num, censored, ch2_tables, ch2_table_name) %>%
  keep_years()

if (!is.null(COMPACT_DOMAINS_KEEP)) {
  compact_county <- compact_county %>% filter(domain_desc %in% COMPACT_DOMAINS_KEEP)
}

compact_county <- compact_county %>%
  drop_other() %>%
  arrange(county_name, ch2_tables, short_desc, year)

write_export(compact_county, FILE_COMPACT_COUNTY,
             "Ch 2 Tables 1-57 compact: Ag Census Areas, Census years, TOTAL + NAICS domains")

cat("Saved:", FILE_COMPACT_COUNTY, "—", nrow(compact_county), "rows |",
    n_distinct(compact_county$county_name), "areas\n")
```

# Quality Control {.tabset}

## Export Manifest

```{r qc_manifest}
bind_rows(export_manifest) %>%
  kable(caption = glue("Partner exports written to {EXPORT_DIR}"),
        col.names = c("File", "Contents", "Rows", "Data items", "Years", "KB"))
```

## Table 1 Coverage

```{r qc_table1_coverage}
t1_all <- bind_rows(t1_state, t1_area)
t1_pub <- t1_all %>% filter(row_type == "NASS Table 1")

if (nrow(t1_pub) > 0) {
  t1_pub %>%
    group_by(geo_name, year) %>%
    summarise(cells = sum(!is.na(value_num)), .groups = "drop") %>%
    pivot_wider(names_from = year, values_from = cells, values_fill = 0, names_sort = TRUE) %>%
    kable(caption = glue("Table 1 cells with a published number, of {nrow(t1_spec)} (suppressed and absent cells excluded)"))
}
```

```{r qc_table1_censored}
if (nrow(t1_pub) > 0) {
  t1_pub %>%
    filter(censored) %>%
    count(geo_name, value, name = "cells") %>%
    pivot_wider(names_from = value, values_from = cells, values_fill = 0) %>%
    kable(caption = "Table 1 cells published as a NASS code instead of a number")
}
```

## Table 1 by NASS Group

```{r qc_table1_groups}
# First-pass triage: Table 1 rows rolled up to NASS's own sector_desc / group_desc
no_cat <- t1_all %>%
  filter(is.na(sector_desc) | is.na(group_desc)) %>%
  distinct(table_row, data_item)
if (nrow(no_cat) > 0) {
  cat("WARNING —", nrow(no_cat), "Table 1 rows have no NASS sector/group\n")
  kable(no_cat, caption = "Table 1 rows with no NASS sector/group")
}

t1_all %>%
  filter(geo_level == "STATE") %>%
  group_by(sector_desc, group_desc) %>%
  mutate(rows = n_distinct(table_row)) %>%
  group_by(sector_desc, group_desc, rows, year) %>%
  summarise(cells = sum(!is.na(value_num)), .groups = "drop") %>%
  pivot_wider(names_from = year, values_from = cells, values_fill = 0, names_sort = TRUE) %>%
  arrange(sector_desc, group_desc) %>%
  kable(caption = glue("Table 1 rows by NASS sector and group, {STATE_ALPHA} state level: rows, then cells with a published number by Census year"))
```

## Table 1 Added Rows

```{r qc_table1_added}
  t1_added <- t1_all %>% filter(!row_type %in% c("NASS Table 1", MV_COPY_ROW_TYPE))

if (nrow(t1_added) > 0) {
  t1_added %>%
    filter(geo_level == "STATE") %>%
    select(table_row, var_name, year, value) %>%
    pivot_wider(names_from = year, values_from = value, names_sort = TRUE) %>%
    arrange(table_row) %>%
    select(-table_row) %>%
    kable(caption = glue("Added rows, {STATE_ALPHA} state level (value or (D))"))
} else {
  cat("No added rows (ADD_CONTEXT_ROWS is FALSE or no total-sales rows were found).\n")
}
```

```{r qc_table1_zero_filled}
filled <- t1_added %>%
  filter(grepl("counted as 0", note)) %>%
  distinct(geo_name, year) %>%
  count(geo_name, name = "years_with_no_food_fish_row")

if (nrow(filled) > 0) {
  kable(filled, caption = "Geography-years with no food fish row, counted as $0 (FOOD_FISH_MISSING_AS_ZERO)")
}
```

```{r qc_table1_missing}
# Table 1 cells with no row for the state in any Census year — for Alaska this is
# mostly commodities not grown here (cotton, rice, peanuts, ...)
t1_spec %>%
  anti_join(t1_state %>% distinct(short_desc, domain_desc, domaincat_desc),
            by = c("short_desc", "domain_desc", "domaincat_desc")) %>%
  select(table_row, data_item, unit_desc) %>%
  kable(caption = glue("Table 1 cells with no {STATE_ALPHA} state-level row in any year"),
        col.names = c("Row", "Data item", "Unit"))
```

## Duplicate Check

```{r qc_duplicates}
dup_t1 <- t1_all %>%
  count(geo_level, geo_name, year, table_row, row_type) %>%
  filter(n > 1)

if (nrow(dup_t1) == 0) {
  cat("OK — one row per geography × year × Table 1 cell × row_type in files (1) and (2).\n")
} else {
  cat("WARNING —", nrow(dup_t1), "geography × year × cell × row_type combinations have more than one row.\n")
  kable(dup_t1, caption = "Duplicated Table 1 keys")
}
```

## API Failures

```{r qc_failures}
failed <- api_log_df %>% filter(status == "failed")

if (nrow(failed) == 0) {
  cat("No failed API calls. Every chunk is cached for the next knit.\n")
} else {
  cat("WARNING:", nrow(failed), "API calls failed and were not cached —",
      "re-knit to retry just these.\n")
  failed %>%
    mutate(error = substr(error, 1, 120)) %>%
    select(geo, call, error) %>%
    kable(caption = "Failed API calls")
}
```

# Session Summary

```{r session_summary}
cat("========================================\n")
cat("AFPC PARTNER EXPORTS — SESSION SUMMARY\n")
cat("========================================\n")
cat("State:", STATE_ALPHA, "\n")
cat("Completed:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
# cat("API pull:", pull_minutes, "minutes |",
#     sum(api_log_df$status == "cached"), "cached /",
#     sum(api_log_df$status != "cached"), "live calls |",
#     sum(api_log_df$status == "failed"), "failed\n")
if (DATA_SOURCE == "saved") {
  cat("Data source: saved master CSVs in", OUT_DIR, "(no API calls)\n")
} else {
  cat("Data source: Quick Stats API,", pull_minutes, "minutes |",
      sum(api_log_df$status == "cached"), "cached /",
      sum(api_log_df$status != "cached"), "live calls |",
      sum(api_log_df$status == "failed"), "failed\n")
}
cat("Exports:", EXPORT_DIR, "\n")
for (m in export_manifest) cat(sprintf("  %-55s %8s rows\n", m$file, format(m$rows, big.mark = ",")))
cat("========================================\n")
```

---

*Report generated `r format(Sys.time(), '%B %d, %Y at %I:%M %p')` — `r CONTACT_NOTE`*
